XRONOS: an open data infrastructure for archaeological chronology

### 2024-12-05 paper.qmd
Add data availability statement and license

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joseph.a.roe@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it easier to find information of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
#published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
number-sections: true
bibliography: 
  - references.bib
  - data/c14_datasets.bib
format:
  elsevier-pdf:
    keep-tex: true    
    cite-method: citeproc
    journal:
      name: Journal of Computer Applications in Archaeology
      formatting: doubleblind
  # Pretty but broken:
  # https://github.com/andrewheiss/hikmah-academic-quarto/issues/7
  # hikmah-pdf:
  #   papersize: a4
  #   mainfont: "Linux Libertine O"
  #   sansfont: "URW Gothic"
  #   biblatex-chicago: true
  #   biblio-style: authordate
  #   biblatexoptions:
  #     # - backend=biber
  #     - autolang=hyphen
  #     - isbn=false
  #     - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
execute:
  echo: false
  cache: true
---

@@

Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is as much organisational than technical.
The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
The databases of the instance at <https://xronos.ch> are also archived with Zenodo (<link omitted for blind review><!-- TODO: link -->).
With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
Equally importantly, we enable the 'right to fork' should others wish to take the software and/or database in another direction.
But aside from these extreme scenarios, the long-term sustainability of XRONOS is contingent on the existence of a community of scholars that use and contribute back to it.

@@

>>>>

---
title: "XRONOS: an open data infrastructure for archaeological chronology"
short-title: XRONOS
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    email: joeroe@hey.com
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [writing, editing, curation, software, analysis]
  - name: Clemens Schmid
    orcid: 0000-0003-3448-5715
    affiliation:
    - ref: mpg
    roles: [curation, software, editing]
  - name: Setareh Ebrahimiabareghi
    orcid: 0000-0003-3749-3147
    affiliation:
    - ref: iaw
    roles: [curation, investigation, editing]
  - name: Caroline Heitz
    orcid: 0000-0001-7188-6775
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, editing]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, writing, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    url: https://www.iaw.unibe.ch/
  - id: mpg
    name: Max Planck Institute for Geoanthropology
    url: https://www.shh.mpg.de/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world. By collating a large number of existing regional and global compilations of dates, XRONOS offers the most comprehensive radiocarbon database yet published, with over 350,000 radiocarbon and 75,000 site records. It also provides a foundation for expanding the systematic collection of chronometric information beyond radiocarbon, with support for typological and dendrochronological dates and a generalisable data model that can be adapted to other methods of absolute dating. Automated and semi-automated quality control processes ensure that data from diverse sources is continuously integrated and standardised, making it easier to find information of interest and reducing the need for manual data cleaning by end users. In this paper we describe the concept and implementation of XRONOS in relation to the state of the art in chronometric data-sharing, and evaluate its potential as a general-purpose open repository and curation platform for archaeological chronology.
keywords: 
  - open data
  - chronology
  - chronometry
  - radiocarbon dating
  - dendrochronology
  - typological dating
#published: "Manuscript for submission to the *Journal of Computer Applications in Archaeology*"
number-sections: true
bibliography: 
  - references.bib
  - data/c14_datasets.bib
format:
  elsevier-pdf:
    keep-tex: true    
    cite-method: citeproc
    journal:
      name: Journal of Computer Applications in Archaeology
      formatting: review
  # Pretty but broken:
  # https://github.com/andrewheiss/hikmah-academic-quarto/issues/7
  # hikmah-pdf:
  #   papersize: a4
  #   mainfont: "Linux Libertine O"
  #   sansfont: "URW Gothic"
  #   biblatex-chicago: true
  #   biblio-style: authordate
  #   biblatexoptions:
  #     # - backend=biber
  #     - autolang=hyphen
  #     - isbn=false
  #     - uniquename=false
#  html:
#    page-layout: article
#    toc: true
#    embed-resources: true
execute:
  echo: false
  cache: true
---

@@

Looking beyond the near term, we have endeavoured to create a sustainable infrastructure that can be maintained by a wider scholarly commons – though we must acknowledge that this is a difficult problem, and one that is as much organisational than technical.
The source code for all the software components of the system are available online (at <https://github.com/xronos-ch>) under open licenses.
The databases of the instance at <https://xronos.ch> are also archived with Zenodo (*link omitted for blind review*<!-- TODO: link -->).
With these two resources we reduce XRONOS' 'bus factor'; if we are not able to continue operating XRONOS, somebody else can fully recreate it.
Equally importantly, we enable the 'right to fork' should others wish to take the software and/or database in another direction.
But aside from these extreme scenarios, the long-term sustainability of XRONOS is contingent on the existence of a community of scholars that use and contribute back to it.

@@

# Data accessibility statement

The data and R code used to produce our analysis of the state of the art in radiocarbonc compilation, including all the figures presented here, can be accessed via Zenodo at (<!-- TODO: add link -->*link omitted for blind review, included in submission materials*)

The database and software described in this paper is open source and can be accessed at <https://github.com/xronos-ch>.

### 2024-12-05 paper.tex
Add data availability statement and license

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  doubleblind]{elsarticle}

@@

\begin{frontmatter}
\title{XRONOS: an open data infrastructure for archaeological
chronology}
\author[1]{Joe Roe%
\corref{cor1}%
}
 \ead{joseph.a.roe@unibe.ch} 
\author[2]{Clemens Schmid%
%
}

@@

Looking beyond the near term, we have endeavoured to create a
sustainable infrastructure that can be maintained by a wider scholarly
commons -- though we must acknowledge that this is a difficult problem,
and one that is as much organisational than technical. The source code
for all the software components of the system are available online (at
\url{https://github.com/xronos-ch}) under open licenses. The databases
of the instance at \url{https://xronos.ch} are also archived with Zenodo
(). With these two resources we reduce XRONOS' `bus factor'; if we are
not able to continue operating XRONOS, somebody else can fully recreate
it. Equally importantly, we enable the `right to fork' should others
wish to take the software and/or database in another direction. But
aside from these extreme scenarios, the long-term sustainability of
XRONOS is contingent on the existence of a community of scholars that
use and contribute back to it.

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  review]{elsarticle}

@@

\begin{frontmatter}
\title{XRONOS: an open data infrastructure for archaeological
chronology}
\author[1]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[2]{Clemens Schmid%
%
}

@@

Looking beyond the near term, we have endeavoured to create a
sustainable infrastructure that can be maintained by a wider scholarly
commons -- though we must acknowledge that this is a difficult problem,
and one that is as much organisational than technical. The source code
for all the software components of the system are available online (at
\url{https://github.com/xronos-ch}) under open licenses. The databases
of the instance at \url{https://xronos.ch} are also archived with Zenodo
(\emph{link omitted for blind review}). With these two resources we
reduce XRONOS' `bus factor'; if we are not able to continue operating
XRONOS, somebody else can fully recreate it. Equally importantly, we
enable the `right to fork' should others wish to take the software
and/or database in another direction. But aside from these extreme
scenarios, the long-term sustainability of XRONOS is contingent on the
existence of a community of scholars that use and contribute back to it.

@@

\section{Data accessibility
statement}\label{data-accessibility-statement}

The data and R code used to produce our analysis of the state of the art
in radiocarbonc compilation, including all the figures presented here,
can be accessed via Zenodo at (\emph{link omitted for blind review,
included in submission materials})

The database and software described in this paper is open source and can
be accessed at \url{https://github.com/xronos-ch}.
